The eco-evolutionary impacts of collective straying on metapopulation dynamics

### 2017-05-27 manuscript/msv3.tex
msv3 and fig:relax changes

%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Lost in space: the eco-evolutionary impacts of straying on metapopulation robustness}
\author{Justin D. Yeakel${}^{1,2,*}$, Jean Philippe Gibert${}^{1}$, Peter Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

@@

%Motivation: straying
Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
Anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) embody these astonishing migrations by homing with high accuracy and precision to their natal streams for reproduction after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}.
Nothwithstanding their philopatric tendancy, not all individuals home, but rather `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf,H:2013fs}.
Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
Most recently the role of social interactions and collective navigation has been hypothesized \cite[][; this volume]{Berdahl:2015kv,Berdahl:2016dx}.
Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.

%The tradeoff
The dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations as well as the evolutionary dynamics of mixed trait distributions that respond to alternative local conditions.
Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of metapopulations (general REF).
For example, the long-term sustainability of the Bristol Bay sockeye salmon fishery is due in large part to intact habitat that in conjunction with fine-scale homing and reduced gene flow provides a template for the evolution of countless locally adapted populations \cite{Hilborn:2003gf,Schindler:2010he,Anonymous:2014ku,Satterthwaite:2015ge}.
In contrast, areas such as California's Central Valley where habitats have been lost, degraded, and homogenized stand in marked contrast.
Within species diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
Weakened portfolio dynamics can also emerge from homogenization resulting from  reduced genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
Of course, these two measures of diversity often occur concurrently: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.

%Eco-evolutionary dynamics in space
%% From geographic mosaic of coevolution to...
That evolutionary forces play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology (REFS).
These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space (REFS).
Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics over contemporary timescales \cite{Hendry:2016un}.
How the ecological dynamics of phenotypically diverse metapopulations, wherein constituent populations are subject to different selective regimes, are altered affected by evolutionary forces is less well known. % and the subject of this paper.

%Movement through space (link back to straying)
Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock in every school or flock (figure \ref{fig:xkcd}).
The rate at which individuals stray, $m$, has been subject to a growing amount of both theoretical and empirical research \cite{H:2013fs,Keefer:2014gg,Bett:2017ha} and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
Regardless of the mechanism's governing straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
Whether, and to what extent, the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selection mosaic is unclear.
How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.

@@

%Alternative stable state
Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
An important and relatively novel component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments \cite[though see ]{Berdahl:2015kv}.
Although our proposed model is was constructed with the dynamics of salmon populations in mind, the framework is general and the conclusions are likely relevant to a diverse range of migratory organisms where locally adapted populations are linked by dispersal.
We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
The emergence of alternative steady states results in a nonlinear response of the portfolio effect as well as the time required for the metapopulation to recover after an induced disturbance, suggesting that metapopulation robustness can be quite sensitive to the combined influences of dispersal across a selective landscape.

%Extinction and return times

%Habitat heterogeneity
A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
However, if we suppose that the rate of straying between two sites is correlated with distance (as indicated by empirical evidence), and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant, suggesting a limited role of collective dispersal on the dynamics considered herein.

%Concluding paragraph of the intro

@@

\noindent{\bf (a) Metapopulation framework}\\
\noindent We consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.

The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.

@@

The combined recruitment of local individuals $(1-m)N_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$.
Given the local trait optimum is $\theta_i$, recruitment is

@@

\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ determines scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
\\

@@

\noindent{\bf (b) Recruitment over a selective landscape}
\noindent Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where 
\begin{equation}
w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
\end{equation}
We make two simplifying assumptions.
First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value being that of the mixed-normal.
Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant.

An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
The mean trait value thus changes through time according to the difference equation

@@

%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
Although there is a direct eignvalue relationship between the rate of return following a small pulse perturbation (REF), because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.

Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
\\

\noindent {\bf (d) The effects of density and distance on the rate of straying}
%Density dependent m
\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the straying rate is density dependent \cite{Berdahl:2016dx,Bett:2017ha}.
Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
According to Berdahl et al. \cite{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is

@@

%Linking m/m0 and thetadiff
The straying rate is intrinsically linked to the distance between the donor and recipient population.
The greater the distance between two populations, the lower the expected rate of straying (REF).
We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.9\textwidth]{figs2/fig_relax_comb.pdf}
\caption{
(a, b) Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming constant straying rates $m$ and (c, d) density dependent straying rates (evaluated at the steady state $m^*$).
If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
} \label{fig:relax}
\end{figure*}

@@

\noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
\noindent Straying generally lowers steady state densities for both populations by \emph{i}) the donor population losing locally-adapted individuals to the recipient population and \emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population, and this accords with observations from natural populations \cite{Bett:2017ha} %, regardless of trait heritability and variance or habitat heterogeneity
The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative steady states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density, which in our system is random due to a small amount of introduced variation in the initial conditions.
The fold bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.

>>>>

%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{The eco-evolutionary impacts of collective straying on metapopulation dynamics}
\author{Justin D. Yeakel${}^{1,2,*}$, Jean Philippe Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

@@

Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size vary across a metapopulation. 
This asynchrony will increase the potential for demographic rescue \cite{Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}. 
For example, different responses to climate variability within populations of a rare plant reduced abundance fluctuations \cite{Abbott:2017hl}. 
This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \cite{Thibaut:2012km}.

Movement of individuals across populations (i.e., straying or dispersal) can have a large influence on metapopulation dynamics  \cite{MilnerGulland:2011vm}. 
First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
Second, high levels of straying may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
Third, straying will influence the evolutionary dynamics of the metapopulation, however evolutionary impacts are less well understood. 
Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}. 
Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.

% %The tradeoff
% Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of metapopulations (general REF).
% For example, the long-term sustainability of the Bristol Bay sockeye salmon fishery is due in large part to intact habitat that in conjunction with fine-scale homing and reduced gene flow provides a template for the evolution of countless locally adapted populations \cite{Hilborn:2003gf,Schindler:2010he,Anonymous:2014ku,Satterthwaite:2015ge}.
% In contrast, areas such as California's Central Valley where habitats have been lost, degraded, and homogenized stand in marked contrast.
% Within species diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
% This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
% Weakened portfolio dynamics can also emerge from homogenization resulting from  reduced genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
% Of course, these two measures of diversity often occur concurrently: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.

%Movement through space (link back to straying)
Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
There is growing appreciation that the rate of straying may be influenced by abiotic, biotic, and anthropogenic factors \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}. 
Recently the role of social interactions and collective navigation has been hypothesized [This issue].
The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
Regardless of the mechanism's governing straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
Whether, and to what extent, the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.

%Eco-evolutionary dynamics in space
%% From geographic mosaic of coevolution to...
That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.
How the ecological dynamics of phenotypically diverse metapopulations, wherein constituent populations are subject to different selective regimes, are altered affected by evolutionary forces is less well known. % and the subject of this paper.

%Motivation: straying
The eco-evolutionary impacts of straying likely has important implications for conservation and management in key taxa such as in salmon.
Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \cite{Schtickzelle:2007wb,Anderson:2014cx}.
Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.  
%Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are reknown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
These strays can introduce new maladaptive genotypes into the recipient population.
Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}. 
Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}. 
%Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.

@@

%Alternative stable state
Here we seek to explore how the potentially detrimental evolutionary effects of straying (erosion of local adaptation) and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
% Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
[We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]

% An important and relatively novel component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments \cite[though see ]{Berdahl:2015kv}.
% Although our proposed model is was constructed with the dynamics of salmon populations in mind, the framework is general and the conclusions are likely relevant to a diverse range of migratory organisms where locally adapted populations are linked by dispersal.
% We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
% The emergence of alternative steady states results in a nonlinear response of the portfolio effect as well as the time required for the metapopulation to recover after an induced disturbance, suggesting that metapopulation robustness can be quite sensitive to the combined influences of dispersal across a selective landscape.
% 
% %Extinction and return times
% 
% %Habitat heterogeneity
% A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
% However, if we suppose that the rate of straying between two sites is correlated with distance (as indicated by empirical evidence), and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
% Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant, suggesting a limited role of collective dispersal on the dynamics considered herein.
% 
% %Concluding paragraph of the intro
% 
%

@@

\noindent{\bf (a) Metapopulation framework}\\
\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determing recruitment rates.
We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.

%PW: Can we say anything about the effect of the recipient population size and that effect on evolutionary dynamics? Again, straying can be thought of as both from the donor or recipient population perspective.
The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.

@@

The recruitment of local individuals $(1-m)N_i(t)$ as a function of their mean trait value at time $t$ and the local trait optimum, is

@@

\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum (the strength of selection increases with smaller values of $\tau$), which we set as $\tau=1$ here and throughout.
Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished, and this matches what is observed in wild populations \cite{Peterson:2014gy}.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
\\

@@

\noindent{\bf (b) Recruitment over a selective landscape}
\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where 
\begin{equation}
w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
\end{equation}
We make two simplifying assumptions.
First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gibert:tz}.

An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
Following Lande \cite{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation

@@

%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \cite{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.

Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of metapopulation fragility.
For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
We measured the time required for the system $N_T$ to recover to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
\\

\noindent {\bf (d) The effects of density and distance on the rate of straying}
%Density dependent m
\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including straying) the straying rate is density dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is

@@

%Linking m/m0 and thetadiff
The straying rate is largely influenced by the distance between the donor and recipient population.
The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
\caption{
Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
} \label{fig:relax}
\end{figure*}

@@

\noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
\noindent Regardless of density dependence, straying lowers steady state densities for both populations by \emph{i}) the donor population losing locally-adapted individuals to the recipient population and \emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig. \ref{fig:traj}).
This prediction is in accordance with observations from natural populations \cite{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative steady states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
At high stray levels, there are two eco-evolutionary states: a population will either have a higher degree of local adaptation (low trait offset) and higher density or be more maladapted and have lower density. 
Whether the population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.

Trait heritability has a large impact on the sensitivity of the straying rate to both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
The fold bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
Although trait heritability among salmonids is variable, most life history traits show $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
